Approximation as Differentiation

### 2025-07-03 main/approx-as-tangents.tex
Fix enumerated list inside example environment.

Our thesis is that \GPS is a form of differentiable programming, where tangents are not linear approximations of functions but instead are qualitiative approximations of elements in the sense of domain theory. Smooth functions in this setting are Berry's \emph{stable functions} \cite{berry79,berry82}. We now introduce these concepts and how they relate to \GPS.

@@

\begin{example}
  \begin{enumerate}
  \item The function $\mathrm{strictOr}$ is stable. For example, for the input-output pair $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$, the minimal input that gives this output is exactly $(\mathsf{tt}, \mathsf{ff})$. If we take the approximation $\bot \leq \mathsf{tt}$ of the output, then the corresponding minimal input is $(\bot, \bot)$. The function $\mathrm{shortCircuitOR}$ is also stable. For the input-output pair $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$, the minimal input that gives this input is $(\mathsf{tt},\bot)$, indicating that the presence of $\mathsf{ff}$ in the second argument was not necessary to produce this output. As with $\mathrm{strictOr}$, the minimal input required to produce the output $\bot \leq \mathsf{tt}$ is again $(\bot,\bot)$.

>>>>

Our thesis is that \GPS is a form of differentiable programming, where tangents are not linear approximations of functions but instead are qualitative approximations of elements in the sense of domain theory. Smooth functions in this setting are Berry's \emph{stable functions} \cite{berry79,berry82}. We now introduce these concepts and how they relate to \GPS.

@@

\begin{example}
  \leavevmode\\[-3ex]
  \begin{enumerate}
  \item The function $\mathrm{strictOr}$ is stable. For example, for the input-output pair $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$, the minimal input that gives this output is exactly $(\mathsf{tt}, \mathsf{ff})$. If we take the approximation $\bot \leq \mathsf{tt}$ of the output, then the corresponding minimal input is $(\bot, \bot)$. The function $\mathrm{shortCircuitOR}$ is also stable. For the input-output pair $(\mathsf{tt},\mathsf{ff}) \mapsto \mathsf{tt}$, the minimal input that gives this input is $(\mathsf{tt},\bot)$, indicating that the presence of $\mathsf{ff}$ in the second argument was not necessary to produce this output. As with $\mathrm{strictOr}$, the minimal input required to produce the output $\bot \leq \mathsf{tt}$ is again $(\bot,\bot)$.

### 2025-07-03 main/related-work.tex
Fix enumerated list inside example environment.

Automatic differentiation (AD), discussed in \secref{first-order:autodiff}, is the idea of computing
derivatives of functions expressed as programs by systematically applying the chain rule. The observation that
these derivative computations could be interleaved with the evaluation of the original program is due to
\citet{linnainmaa76}, who showed how the forward derivative $\pushf{f}_x$ of $f$ at a point $x$ could be
computed alongside $f(x)$ in a single pass, dramatically improving the efficiency of derivative evaluation
over symbolic or numerical differentiation. This insight became the foundation of forward-mode AD, which
underpins many optimisation and scientific computing tools, including JAX~\cite{jax2018github}.

More recent approaches to automatic differentiation have emphasised its semantic foundations.
\citet{elliott18} proposed a categorical model of AD that interprets programs as functions enriched with their
derivatives, giving a compositional account of differentiation based on duality and linear maps. This line of
work connects AD to denotational semantics, abstracting over operational concerns and enabling principled
extensions to differentiable programming languages. Vákár and collaborators~\cite{vákár22,nunes2023} developed
the CHAD framework upon which this paper was based, using Grothendieck constructions over indexed categories
to capture both values and their tangents in a compositional semantic structure. These semantic perspectives
shed light on the categorical structure of AD and guide the design of systems that generalise beyond real
analysis, including the application to data provenance and slicing explored in this paper.

>>>>

Automatic differentiation (AD), discussed in \secref{first-order:autodiff}, is the idea of computing
derivatives of functions expressed as programs by systematically applying the chain rule. The observation that
these derivative computations could be interleaved with the evaluation of the original program is due to
\citet{linnainmaa76}, who showed how the forward derivative $\pushf{f}_x$ of $f$ at a point $x$ could be
computed alongside $f(x)$ in a single pass, dramatically improving the efficiency of derivative evaluation
over symbolic or numerical differentiation. This insight became the foundation of forward-mode AD, which
underpins many optimisation and scientific computing tools, including JAX~\cite{jax2018github}.
\citet{griewank89} showed how the Wengert list, the linear record of assignments used in forward-mode to
compute derivatives efficiently, could be traversed in reverse to compute the pullback map. This two-pass
approach is the foundation of reverse-mode AD, and closely resembles implementations of Galois slicing
(\secref{related-work:galois-slicing} below) that record a trace during forward slicing for use in backward
slicing.

More recent approaches to automatic differentiation have emphasised semantic foundations. \citet{elliott18}
proposed a categorical model of AD that interprets programs as functions enriched with their derivatives,
giving a compositional account of differentiation based on duality and linear maps. Vákár and
collaborators~\cite{vákár22,nunes2023} developed the CHAD framework which inspired this paper, using
Grothendieck constructions over indexed categories to capture both values and their tangents in a
compositional semantic structure. These perspectives shed light on the categorical structure of AD and guide
the design of systems that generalise AD beyond real analysis, including the application to data provenance
and slicing explored in this paper.
